### 2022-07-23 src/ms.tex
Add section on polynomial root solver

% Custom command
\newcommand{\ud}{\,\mathrm{d}}
\renewcommand{\vec}[1]{\boldsymbol{\mathbf{#1}}}

\hypersetup{
    colorlinks=true,
    allcolors=cyan,
}

@@

Scientific progress in modern astronomy research commonly relies on gathering
large quantities of data using exceedingly precise instruments. The process
which ``generates'' these data consists of the physical phenomenon of interest
-- for instance, an exoplanet blocking or twisting the light of a distant star,
and the noise introduced by the measurement process and the presence of an
atmosphere The task for an astronomer is to first construct a \emph{model}
which describes the entire process which generated the data and to then ``fit''
that model to data. All models only approximate reality and the researcher has
to make a series of decisions during the model building process, everything
from how to process the raw data to which results to put in the abstract of a
paper. Advancements in computational statistics and machine learning in the
past decade or so have made it possible to fit ever more complex models to
data. These models are generally expressed in computer code which may contain
complex numerical algorithms, such as iterative solvers and numerical
integrals. In this thesis, I mostly focused on developing methods which enable
\emph{statistical inference} with these kinds of complex models in two
particular domains within astronomy, gravitational microlensing and occultation
mapping. The common theme between these two topics is that both deal with
accurately measuring the brightness of distant stars as a function of time with
the goal of inferring properties of exoplanets and stars. Broadly speaking, I
believe the biggest contribution of this thesis is providing a new lens for
looking at a particular set of old problems, a lens which incorporates recent
advancements from statistics, machine learning and computer science. More
specifically, I have developed a an open-source software package
\textsf{caustics}\footnote{\url{https://github.com/fbartolic/caustics}} which
enables fast and accurate computation of binary lens and triple lens
microlensing light curves and simultaneously provides exact \emph{gradients} of
the code outputs with respect to all its inputs. This is significant because it
for the first time enables the use of modern gradient based statistical
inference algorithms such as Hamiltonian Monte Carlo with microlensing light
curves. Microlensing is one of the major goals for the upcoming \emph{NASA
    Roman} telescope and the existing modeling methods are completely inadequate
for dealing with the scale of data which will come from \emph{Roman}. I also
propose a framework for dealing with issues which have plagued the field for
decades -- various pathologies in microlensing models and questions about the
interpretation of statistical results. Besides microlensing, I have also delved
into the field of occultation mapping of Solar System objects and exoplanets.
Together with collaborators, I have developed a novel statistical method for
reconstructing spatial maps of volcanic emission on Jupiter's moon Io using
infra-red occultation light curves. I applied the same method to exoplanets to
explore the exciting possibility of detecting weather changes on Hot Jupiters
by reconstructing two dimensional maps of the emission surface from simulated
\emph{JWST} secondary eclipse light curves. I found that planetary scale
changes in the emission pattern should be detectable with \emph{JWST}.

@@

The assumptions that existing methods for modeling microlensing events rely on
are often opaque and unquestioned. Discussions on model ``degeneracies''
\citep{2014MNRAS.437.4006S,2019AJ....157...23H,2018AcA....68...43S,2009MNRAS.393..816D},
correlated noise \citep{2015ApJ...812..136B,2019MNRAS.488.3308L} and model
comparison \citep{2018AJ....155..259H,2019MNRAS.484.5608D} have been ongoing in
the microlensing literature for decades without a clear solution and proper
framing of the issue. In this thesis I will revisit these kinds of questions
while taking into account many recent developments in the fields of
computational statistics and machine learning. My main contributions are the
following:
\begin{itemize}
    \item I wrote an open-source Python package \textsf{caustics} which enables fast and
          accurate computation of binary and triple lens microlensing light curves using
          contour integration. The code runs on both CPU and GPU architectures, and
          crucially, for reasons that I will elaborate in subsequent chapters, it
          supports \emph{automatic differentiation} of all outputs with respect to all
          input parameters which enables the use of statistical methods which are orders
          of magnitude more efficient than existing approaches. The code can easily be
          extended to support quadruple lensing and arbitrary intensity profiles for the
          source star.
    \item I have revisited the topic of degeneracies in the posterior probability
          distributions which appear in the context of microlensing and propose solutions
          to these problems.
    \item I have extended the complex polynomial root solver from \citet{Cameron2021}
          such that it can be executed on GPU architectures and that it supports
          automatic differentiation. This enables the evaluation of ~1M lens equation
          solutions for point source binary and triple lenses in seconds using a CPU and
          miliseconds using a GPU.
    \item I propose an approach to statistical inference on a population of microlensing
          events which different from existing methods and has some advantages.
    \item I briefly discuss the issue of correlated noise in microlensing light curves
          and propose ways to account for it using Gaussian Processes.
\end{itemize}

@@

In recent years there have been significant advances in statistical modeling of
phase curves and eclipse light curves. Most notably,
\citet{2019AJ....157...64L} introduced the \textsf{starry} code which enables
analytic computation of phase curves and occultation light curves for bodies
with arbitrary emission maps expressed in a spherical harmonic basis (an idea
dating back to \citet{1906ApJ....24....1R} ) and \citet{2021arXiv210306275L}
expanded the algorithm for the (considerably more complicated) case of
reflected light. In this thesis, I will present the work I've done in
collaboration with other researchers from the planetary science and exoplanet
communities on using \textsf{starry} to map the surface of Io and investigate
the prospects for detecting fine spatial structure in the atmospheres of Hot
Jupiters using JWST phase curves and eclipse light curves. My contributions to
this area are the following:
\begin{itemize}
    \item I have developed a novel model for inferring emission maps of Io from
          occultation light curves. First part of this project is published in
          \citet{2022PSJ.....3...67B} though I haven't managed to complete the second
          part.
    \item I have investigated the problem of eclipse mapping of exoplanets using the
          \textsf{starry} framework, focusing in particular on the possibility of
          detecting planetary scale storms on Hot Jupiters using JWST. I found that it
          extremely difficult to infer maps of higher resolution than a dipole order but
          that should still be sufficient for detecting large scale weather and climate
          change in the atmospheres of these planets.
\end{itemize}
The application of the occultation/eclipse mapping method to Io can be seen as the
best case scenario for the application of the same method to exoplanets.
I will show that even in the case of observing a bright object right in our neighborhood,
this is by no means and easy task. Our ability to reconstruct the surface features of
Io sets a sort of upper limit on what is possible with exoplanets.

%\section{Outline}
%The outline of the thesis is as follows. Chapter~\ref{ch:theoretical_min}
%contains the theoretical background needed to understand the subsequent
%chapters, covering the theory behind microlensing and occultation mapping, a
%summary of Bayesian statistics in particular and statistical inference more
%generally, and probabilistic programming which ties everything together.
%Chapter~\ref{ch:microlensing} is the most extensive chapter in the thesis
%covering all of my work on microlensing. I start by discussing the case of
%single lens microlensing models, showing that it contains many of the
%properties of more complex models. In particular I focus on the issue of
%multi-modal posterior distributions which is a ubiquitous problem in
%microlensing. I propose several solutions to this and demonstrate their
%effectiveness. I then move on to the considerably more complex case of binary
%and triple lens models starting with numerical methods for solving the lens
%equation (finding roots of a complex polynomial equation). I demonstrate the
%advantage of the Ehrlich-Aberth algorithm over previous methods and explain how
%to make this algorithm differentiable and orders of magnitude faster through
%the use of GPUs. Next, I cover the contour integration algorithm for computing
%the magnification of a limb-darkened source star and its implementation in
%\textsf{caustics}.

@@

We can rewrite Equation~\ref{eq:lens_equation} by defining new dimensionless
angular coordinates scaled by the angular Einstein radius:
\begin{equation}
    x= \frac{\theta}{D_L\theta_E}, \quad y=\frac{\beta}{D_S\theta_E}
\end{equation}
The lens equation then takes the very simple form
\begin{equation}
    y= x- \frac{ 1}{x}
    \label{eq:lens_equation_dimensionless}
\end{equation}
Equation~\ref{eq:lens_equation_dimensionless} is a nonlinear mapping between the
source plane $y$ and the image plane $x$.
If we solve for $x$, we obtain two solutions for the position of the images, given by

\begin{equation}
    x_\pm= \frac{1}{2} \left( y\pm\sqrt{y^2 +4}\right)
    \label{eq:images_location}
\end{equation}
For non zero $y$ the so called \emph{minor image} is always located
within the Einstein radius ($\lvert x_-\rvert<1$) while the \emph{major image}
is located outside of it ($\lvert x_+\rvert>1$).

Since gravitational lensing does not involve any additional emission or
absorption along a deflected ray of light and since the wavelength of the light
ray does not change when it encounters the lens, the \emph{surface brightness}
(flux density per unit angular area) $I$ of the source image is identical to
the surface brightness of the unlensed source. The flux of an infinitesimal
source is a product of the surface brightness and the solid angle $\Delta
    \omega$ subtended by the source in the sky which changes throughout the
microlensing event because the source and the lens are in motion relative to an
observer. The ratio of the original and the lensed source flux, the
\emph{magnification} $A$, is then given by the ratio of the two solid angles
\begin{equation}
    A= \frac{\Delta \omega}{(\Delta\omega)_0}\quad,
    \label{eq:magnification}
\end{equation}
where $0$ denotes the unlensed solid angle.
Assuming an infinitesimal source size at angular location $\mathbf y =(y_1,y_2)$ subtending an angle
$\Delta (\omega)_0$, and an image of the source at location $\mathbf x=(x_1,x_2)$ subtending an angle
$\Delta \omega$; the ratio between the solid angle of the source and the image
is given by the determinant of the Jacobian matrix of the
lens mapping $\mathbf{x}\rightarrow\mathbf{y}$, evaluated at the location of the images
\begin{equation}
    \frac{(\Delta\omega)_0}{\Delta\omega} =\left\lvert\textrm{det}
    \frac{\partial \mathbf y}{\partial \mathbf x} \right\rvert
\end{equation}
The magnification $A$ is then given by the inverse Jacobian determinant of the lens mapping:
\begin{equation}
    A= \left\lvert\det
    \frac{\partial \mathbf y}{\partial \mathbf x} \right\rvert^{-1} \label{eq:magnification_general}
\end{equation}
The images for which the determinant of the Jacobian of the lens mapping is
positive are said to have positive \emph{parity} and vice versa.

Looking at Equation~\ref{eq:magnification_general}, we see that the
magnification diverges when the Jacobian determinant of the lens mapping
vanishes. Curves in the lens plane for which the determinant of the lens
mapping vanishes, that is,
\begin{equation}
    \left\lvert\det
    \frac{\partial \mathbf y}{\partial \mathbf x} \right\rvert=0
\end{equation}
are called \emph{critical curves}.
The critical curves in the lens plane can be mapped to the source plane using
the lens equation $\mathbf{x}\rightarrow \mathbf{y}$ and these curves are
are called \emph{caustic curves}.
Although the magnification factor in Equation~\ref{eq:magnification_general} formally
diverges at the points of critical curves or caustics,
the divergence is not physical because light sources aren't truly point like.
If the finite angular size of the source is taken into
account, the magnification is an integral of Equation~\ref{eq:magnification_general} over the
extent of the source, weighted by the source brightness and that integral is always finite.
Both the critical and the caustic curves are closed and one can prove that \textbf{the
    number of images changes by two if and only if the source crosses a caustic curve}
\citep[][Chapter 6]{1992grle.book.....S}.

We can derive a simpler expression for Equation~\ref{eq:magnification} by
switching to polar coordinates $(v,\phi)$ in the lens plane. The lens equation
for the two vector components of the apparent source position is given by
\begin{align}
    y_1= & x_1- \frac{x_1}{x_1^2+x_2^2} \\
    y_2= & x_2- \frac{x_2}{x_1^2+x_2^2}
\end{align}
Introducing polar coordinates $x_1=v\cos\phi,\;x_2=v\sin\phi$, we have
\begin{equation}
    \left\lvert\textrm{det}
    \frac{\partial \mathbf y}{\partial \mathbf x} \right\rvert =1- \frac{1}{v^4}\quad,
\end{equation}
By substituting in the locations of the source images Equation~\ref{eq:images_location} into the
above expression, we obtain the magnification of the source for both images
\begin{equation}
    A_\pm=\left(1- \frac{1}{x^4_\pm} \right)^{-1}
\end{equation}
The total magnification is then the sum of the two magnifications, and is given by
\citep{1936Sci....84..506E}
\begin{equation}
    A(u)=\lvert A_-\rvert+\lvert A_+\rvert= \frac{u^2+2}{u\sqrt{u^2+4}}\quad,
    \label{eq:magnification_point_lens}
\end{equation}
where $u = \sqrt{y_1^2+y_2^2}$ is the the magnitude of the position
vector of the source star.
The critical curve of the point lens is the Einstein
ring corresponding to $v=\sqrt{x_1^2+x_2^2}=1$ and the caustic curve is mapped to a
a single point in the source plane at $u=0$.
For source separations much smaller than the angular Einstein radius ($u\ll 1$), the
magnification is approximately $A(u)\simeq 1/u$, in the opposite case ($u\gg 1$)
we have $A(u)\simeq 1+ 2/u^4$, that is, the magnification falls off rapidly the further
away the source is from the lens.

To be able to evaluate Equation~\ref{eq:magnification_point_lens} in practice
we have to parametrize the position of the source on the sky $u$ as a function
of time. To derive an expression for $u(t)$ we assume the the motion of the
observer, lens and the source is rectilinear (acceleration can be neglected).
We take $\vec{x}_S$ to be the angular position of the source star on the plane
of the sky and $\boldsymbol{\mu}_L$ to be its proper motion vector. Likewise
for the lens. We thus have
\begin{align}
    \vec{x}_S(t) & =\vec{x}_{S, 0}+
    \left(t-t_{0}\right) \boldsymbol{\mu}_S
    \label{eq:source_position}      \\
    \vec{x}_L(t) & =\vec{x}_{L, 0}
    +\left(t-t_{0}\right) \boldsymbol{\mu}_L\quad,
    \label{eq:lens_position}
\end{align}
where $t_0$ is some reference time.
The relative position vector of the lens with respect to the source is then
\begin{equation}
    \boldsymbol{u}(t) \equiv \frac{\vec{x}_L(t)
        -\vec{x}_S(t)}{\theta_E}=
    \frac{\vec{x}_{LS, 0}}{\theta_E}
    +\frac{t-t_{0}}{\theta_E} \boldsymbol{\mu}_{LS}
    \label{eq:relative_trajectory_no_parallax}
\end{equation}
where
$\vec{x}_{L S, 0}\equiv\vec{x}_{L,0}-\vec{x}_{S, 0}$
is the relative position at $t_0$ and
$\boldsymbol{\mu}_{LS}\equiv \boldsymbol{\mu}_{L}- \boldsymbol{\mu}_{S}$ is the
relative proper motion.
Since $\boldsymbol{\mu}_{LS}$ and $\vec{x}_{L S, 0}$ are
perpendicular to each other, it follows that the magnitude of the relative
separation is
\begin{equation}
    u(t)=\sqrt{u_0^2+ \left(\frac{t-t_0}{t_E}\right)^2}
\end{equation}
where we have defined $u_0\equiv |\vec{x}_{L S, 0}|/\theta_E$
and $t_E\equiv \theta_E/|\boldsymbol{\mu}_{LS}|$.
The magnification is then a function of three parameters $(t_0, u_0, t_E)$
and the resulting curve as a function of time is often called the Paczy\'nski curve
\citep{1986ApJ...304....1P,1986ApJ...301..503P}.
The magnification as a function of time for different impact parameters $u_0$
is shown in Figure~\ref{fig:paczynski_curve}. Notice the very steep fall off in
magnification as as we move away from the Einstein ring ($A(u)\propto
    1+2u^{-4}$).
In the limits of $u_0\rightarrow 0$ and $u_0\gg 1$  there is a continuous
mathematical degeneracy between the parameters $t_0$, $u_0$ and $t_E$
\citep{1997ApJ...487...55W}.

@@

\subsection{Observed flux}
The magnification $A(t)$ is not a direct observable in microlensing events. We
can only measure the flux of the source star which is usually contaminated with
the flux of other nearby stars in the dense stellar field and potentially also
with light from the lens and possible companions to the lens. The observed flux
is thus given by
\begin{equation}
    F(t)=F_S(t)+F_B(t)
    \label{eq:flux_observed}
\end{equation}
where $F_s$ is the flux of the source star and $F_B$ is the contamination flux also
called the \emph{blending flux}. To quantify the strength of the blending e can
define the emph{source flux fraction} $b_S$ as the fraction of total (unmagnified)
flux that is coming from the source star:
\begin{equation}
    b_S\equiv F_S/(F_S + F_B)
\end{equation}
For highly blended events $b_S\approx 0$ and in absence of blending $b_S=1$.
The source flux $F_S$ and the blending flux $F_B$ are generally highly correlated
and it is often preferable to use a slightly different parametrization proposed by
\citet{2009MNRAS.393..816D}. In this parametrization we use the
\emph{baseline flux} $F_\mathrm{base}$ and the difference between
the baseline flux and the flux at peak magnification $\Delta F\equiv F_S[A(u_0)-1]$
as the new parameters.
Equation~\ref{eq:flux_observed} then takes the form
\begin{equation}
    f=\Delta F\frac{A(t) - 1}{A(t_0)-1}+F_\mathrm{base}
    \label{eq:flux_dominik}
\end{equation}
It is generally straightforward to measure the flux difference $\Delta F$, the
baseline flux $F_\mathrm{base}$ and the center of the curve $t_0$ but getting a
good estimate of $t_E$ even outside of the regime $u_0\rightarrow 0$ and
$u_0\gg 1$ requires good photometric sampling along the ``wings'' of the light
curve \citep{2009MNRAS.393..816D}.

>>>>

% Custom command
\newcommand{\ud}{\,\mathrm{d}}
\newcommand{\ssf}[1]{\textsf{#1}}
\renewcommand{\vec}[1]{\boldsymbol{\mathbf{#1}}}
\newcommand{\hquad}{~~}

@@

Scientific progress in modern astronomy research commonly relies on gathering
large quantities of data using exceedingly precise instruments. The process
which ``generates'' these data consists of the physical phenomenon of interest
-- for instance, an exoplanet blocking or twisting the light of a distant star,
and the noise introduced by the measurement process and the presence of an
atmosphere The task for an astronomer is to first construct a \emph{model}
which describes the entire process which generated the data and to then ``fit''
that model to data. All models only approximate reality and the researcher has
to make a series of decisions during the model building process, everything
from how to process the raw data to which results to put in the abstract of a
paper. Advancements in computational statistics and machine learning in the
past decade or so have made it possible to fit ever more complex models to
data. These models are generally expressed in computer code which may contain
complex numerical algorithms, such as iterative solvers and numerical
integrals. In this thesis, I mostly focused on developing methods which enable
\emph{statistical inference} with these kinds of complex models in two
particular domains within astronomy, gravitational microlensing and occultation
mapping. The common theme between these two topics is that both deal with
accurately measuring the brightness of distant stars as a function of time with
the goal of inferring properties of exoplanets and stars. Broadly speaking, I
believe the biggest contribution of this thesis is providing a new lens for
looking at a particular set of old problems, a lens which incorporates recent
advancements from statistics, machine learning and computer science. More
specifically, I have developed a an open-source software package
\ssf{caustics}\footnote{\url{https://github.com/fbartolic/caustics}} which
enables fast and accurate computation of binary lens and triple lens
microlensing light curves and simultaneously provides exact \emph{gradients} of
the code outputs with respect to all its inputs. This is significant because it
for the first time enables the use of modern gradient based statistical
inference algorithms such as Hamiltonian Monte Carlo with microlensing light
curves. Microlensing is one of the major goals for the upcoming \emph{NASA
    Roman} telescope and the existing modeling methods are completely inadequate
for dealing with the scale of data which will come from \emph{Roman}. I also
propose a framework for dealing with issues which have plagued the field for
decades -- various pathologies in microlensing models and questions about the
interpretation of statistical results. Besides microlensing, I have also delved
into the field of occultation mapping of Solar System objects and exoplanets.
Together with collaborators, I have developed a novel statistical method for
reconstructing spatial maps of volcanic emission on Jupiter's moon Io using
infra-red occultation light curves. I applied the same method to exoplanets to
explore the exciting possibility of detecting weather changes on Hot Jupiters
by reconstructing two dimensional maps of the emission surface from simulated
\emph{JWST} secondary eclipse light curves. I found that planetary scale
changes in the emission pattern should be detectable with \emph{JWST}.

@@

The assumptions that existing methods for modeling microlensing events rely on
are often opaque and unquestioned. Discussions on model ``degeneracies''
\citep{2014MNRAS.437.4006S,2019AJ....157...23H,2018AcA....68...43S,2009MNRAS.393..816D},
correlated noise \citep{2015ApJ...812..136B,2019MNRAS.488.3308L} and model
comparison \citep{2018AJ....155..259H,2019MNRAS.484.5608D} have been ongoing in
the microlensing literature for decades without a clear solution and proper
framing of the issue. In this thesis I will revisit these kinds of questions
while taking into account many recent developments in the fields of
computational statistics and machine learning. My main contributions are the
following:
\begin{itemize}
    \item I wrote an open-source Python package \ssf{caustics} which enables fast and
          accurate computation of binary and triple lens microlensing light curves using
          contour integration. The code runs on both CPU and GPU architectures, and
          crucially, for reasons that I will elaborate in subsequent chapters, it
          supports \emph{automatic differentiation} of all outputs with respect to all
          input parameters which enables the use of statistical methods which are orders
          of magnitude more efficient than existing approaches. The code can easily be
          extended to support quadruple lensing and arbitrary intensity profiles for the
          source star.
    \item I have revisited the topic of degeneracies in the posterior probability
          distributions which appear in the context of microlensing and propose solutions
          to these problems.
    \item I have extended the complex polynomial root solver from \citet{Cameron2021}
          such that it can be executed on GPU architectures and that it supports
          automatic differentiation. This enables the evaluation of ~1M lens equation
          solutions for point source binary and triple lenses in seconds using a CPU and
          miliseconds using a GPU.
    \item I propose an approach to statistical inference on a population of microlensing
          events which different from existing methods and has some advantages.
    \item I briefly discuss the issue of correlated noise in microlensing light curves
          and propose ways to account for it using Gaussian processes.
\end{itemize}

@@

In recent years there have been significant advances in statistical modeling of
phase curves and eclipse light curves. Most notably,
\citet{2019AJ....157...64L} introduced the \ssf{starry} code which enables
analytic computation of phase curves and occultation light curves for bodies
with arbitrary emission maps expressed in a spherical harmonic basis (an idea
dating back to \citet{1906ApJ....24....1R} ) and \citet{2021arXiv210306275L}
expanded the algorithm for the (considerably more complicated) case of
reflected light. In this thesis, I will present the work I've done in
collaboration with other researchers from the planetary science and exoplanet
communities on using \ssf{starry} to map the surface of Io and investigate the
prospects for detecting fine spatial structure in the atmospheres of Hot
Jupiters using JWST phase curves and eclipse light curves. My contributions to
this area are the following:
\begin{itemize}
    \item I have developed a novel model for inferring emission maps of Io from
          occultation light curves. First part of this project is published in
          \citet{2022PSJ.....3...67B} though I haven't managed to complete the second
          part.
    \item I have investigated the problem of eclipse mapping of exoplanets using the
          \ssf{starry} framework, focusing in particular on the possibility of detecting
          planetary scale storms on Hot Jupiters using JWST. I found that it extremely
          difficult to infer maps of higher resolution than a dipole order but that
          should still be sufficient for detecting large scale weather and climate change
          in the atmospheres of these planets.
\end{itemize}
The application of the occultation/eclipse mapping method to Io can be seen as the
best case scenario for the application of the same method to exoplanets.
I will show that even in the case of observing a bright object right in our neighborhood,
this is by no means and easy task. Our ability to reconstruct the surface features of
Io sets a sort of upper limit on what is possible with exoplanets.

%\section{Outline}
%The outline of the thesis is as follows. Chapter~\ref{ch:theoretical_min}
%contains the theoretical background needed to understand the subsequent
%chapters, covering the theory behind microlensing and occultation mapping, a
%summary of Bayesian statistics in particular and statistical inference more
%generally, and probabilistic programming which ties everything together.
%Chapter~\ref{ch:microlensing_modeling} is the most extensive chapter in the thesis
%covering all of my work on microlensing. I start by discussing the case of
%single lens microlensing models, showing that it contains many of the
%properties of more complex models. In particular I focus on the issue of
%multi-modal posterior distributions which is a ubiquitous problem in
%microlensing. I propose several solutions to this and demonstrate their
%effectiveness. I then move on to the considerably more complex case of binary
%and triple lens models starting with numerical methods for solving the lens
%equation (finding roots of a complex polynomial equation). I demonstrate the
%advantage of the Ehrlich-Aberth algorithm over previous methods and explain how
%to make this algorithm differentiable and orders of magnitude faster through
%the use of GPUs. Next, I cover the contour integration algorithm for computing
%the magnification of a limb-darkened source star and its implementation in
%\ssf{caustics}.

@@

We can rewrite Equation~\ref{eq:lens_equation} by defining new dimensionless
angular coordinates scaled by the angular Einstein radius:
\begin{equation}
    z\equiv \frac{\theta}{D_L\theta_E}, \quad w\equiv\frac{\beta}{D_S\theta_E}
\end{equation}
The lens equation then takes the very simple form
\begin{equation}
    z= w- \frac{ 1}{w}
    \label{eq:lens_equation_dimensionless}
\end{equation}
Equation~\ref{eq:lens_equation_dimensionless} is a nonlinear mapping between the
source plane $w$ and the image plane $z$.
If we solve for $z$, we obtain two solutions for the position of the images, given by

\begin{equation}
    z_\pm= \frac{1}{2} \left( w\pm\sqrt{w^2 +4}\right)
    \label{eq:images_location}
\end{equation}
For non zero $w$ the so called \emph{minor image} is always located
within the Einstein radius ($\lvert z_-\rvert<1$) while the \emph{major image}
is located outside of it ($\lvert z_+\rvert>1$).

Since gravitational lensing does not involve any additional emission or
absorption along a deflected ray of light and since the wavelength of the light
ray does not change when it encounters the lens, the \emph{surface brightness}
(flux density per unit angular area) $I$ of the source image is identical to
the surface brightness of the unlensed source. The flux of an infinitesimal
source is a product of the surface brightness and the solid angle $\Delta
    \omega$ subtended by the source in the sky which changes throughout the
microlensing event because the source and the lens are in motion relative to an
observer. The ratio of the original and the lensed source flux, the
\emph{magnification} $A$, is then given by the ratio of the two solid angles
\begin{equation}
    A= \frac{\Delta \omega}{(\Delta\omega)_0}\quad,
    \label{eq:magnification}
\end{equation}
where $0$ denotes the unlensed solid angle.
Assuming an infinitesimal source size at angular location
$\mathbf w \equiv (w_1,w_2)$ subtending an angle
$\Delta (\omega)_0$, and an image of the source at location
$\mathbf z\equiv(z_1,z_2)$ subtending an angle
$\Delta \omega$; the ratio between the solid angle of the source and the image
is given by the determinant of the Jacobian matrix of the
lens mapping $\mathbf{z}\rightarrow\mathbf{w}$, evaluated at the location of
the images
\begin{equation}
    \frac{(\Delta\omega)_0}{\Delta\omega} =\left\lvert\textrm{det}
    \frac{\partial \mathbf w}{\partial \mathbf z} \right\rvert
\end{equation}
The magnification $A$ is then given by the inverse Jacobian determinant
of the lens mapping:
\begin{equation}
    A= \left\lvert\det
    \frac{\partial \mathbf w}{\partial \mathbf z} \right\rvert^{-1} \label{eq:magnification_general}
\end{equation}
The images for which the determinant of the Jacobian of the lens mapping is
positive are said to have positive \emph{parity} and vice versa.

Looking at Equation~\ref{eq:magnification_general}, we see that the
magnification diverges when the Jacobian determinant of the lens mapping
vanishes. Curves in the lens plane for which the determinant of the lens
mapping vanishes, that is,
\begin{equation}
    \left\lvert\det
    \frac{\partial \mathbf w}{\partial \mathbf z} \right\rvert=0
\end{equation}
are called \emph{critical curves}.
The critical curves in the lens plane can be mapped to the source plane using
the lens equation $\mathbf{z}\rightarrow \mathbf{w}$ and these curves are
are called \emph{caustic curves}.
Although the magnification factor in Equation~\ref{eq:magnification_general} formally
diverges at the points of critical curves or caustics,
the divergence is not physical because light sources aren't truly point like.
If the finite angular size of the source is taken into
account, the magnification is an integral of Equation~\ref{eq:magnification_general} over the
extent of the source, weighted by the source brightness and that integral is always finite.
Both the critical and the caustic curves are closed and one can prove that \textbf{the
    number of images changes by two if and only if the source crosses a caustic curve}
\citep[][Chapter 6]{1992grle.book.....S}.

We can derive a simpler expression for Equation~\ref{eq:magnification} by
switching to polar coordinates $(v,\phi)$ in the lens plane. The lens equation
for the two vector components of the apparent source position is given by
\begin{align}
    w_1= & z_1- \frac{z_1}{z_1^2+z_2^2} \\
    w_2= & z_2- \frac{z_2}{z_1^2+z_2^2}
\end{align}
Introducing polar coordinates $z_1=v\cos\phi,\;z_2=v\sin\phi$, we have
\begin{equation}
    \left\lvert\textrm{det}
    \frac{\partial \mathbf w}{\partial \mathbf z} \right\rvert =1- \frac{1}{v^4}\quad,
\end{equation}
By substituting in the locations of the source images
Equation~\ref{eq:images_location} into the
above expression, we obtain the magnification of the source for both images
\begin{equation}
    A_\pm = \frac{1}{2}\left(\frac{u^{2}+2}{u \sqrt{u^{2}+4}} \pm 1\right)
    \label{eq:magnification_per_image}
\end{equation}
where $u = \sqrt{w_1^2+w_2^2}$ is the the magnitude of the position
vector of the source star.
The total magnification is then the sum of the two magnifications, and is given by
\citep{1936Sci....84..506E}
\begin{equation}
    A(u)=\lvert A_-\rvert+\lvert A_+\rvert= \frac{u^2+2}{u\sqrt{u^2+4}}\quad,
    \label{eq:magnification_point_lens}
\end{equation}
The critical curve of the point lens is the Einstein
ring corresponding to $v=\sqrt{z_1^2+z_2^2}=1$ and the caustic curve is mapped to a
a single point in the source plane at $u=0$.
For source separations much smaller than the angular Einstein radius ($u\ll 1$), the
magnification is approximately $A(u)\simeq 1/u$, in the opposite case ($u\gg 1$)
we have $A(u)\simeq 1+ 2/u^4$, that is, the magnification falls off rapidly the further
away the source is from the lens.

To be able to evaluate Equation~\ref{eq:magnification_point_lens} in practice
we have to parametrize the position of the source on the sky $u$ as a function
of time. To derive an expression for $u(t)$ we assume the the motion of the
observer, lens and the source is rectilinear (acceleration can be neglected).
We take $\vec{w}_S$ to be the angular position of the source star on the plane
of the sky and $\boldsymbol{\mu}_L$ to be its proper motion vector. Likewise
for the lens. We thus have
\begin{align}
    \vec{w}_S(t) & =\vec{w}_{S, 0}+
    \left(t-t_{0}\right) \boldsymbol{\mu}_S
    \label{eq:source_position}      \\
    \vec{w}_L(t) & =\vec{w}_{L, 0}
    +\left(t-t_{0}\right) \boldsymbol{\mu}_L\quad,
    \label{eq:lens_position}
\end{align}
where $t_0$ is some reference time.
The relative position vector of the lens with respect to the source is then
\begin{equation}
    \boldsymbol{u}(t) \equiv \frac{\vec{w}_L(t)
        -\vec{w}_S(t)}{\theta_E}=
    \frac{\vec{w}_{LS, 0}}{\theta_E}
    +\frac{t-t_{0}}{\theta_E} \boldsymbol{\mu}_{LS}
    \label{eq:relative_trajectory_no_parallax}
\end{equation}
where
$\vec{w}_{L S, 0}\equiv\vec{w}_{L,0}-\vec{w}_{S, 0}$
is the relative position at $t_0$ and
$\boldsymbol{\mu}_{LS}\equiv \boldsymbol{\mu}_{L}- \boldsymbol{\mu}_{S}$ is the
relative proper motion.
Since $\boldsymbol{\mu}_{LS}$ and $\vec{w}_{L S, 0}$ are
perpendicular to each other, it follows that the magnitude of the relative
separation is
\begin{equation}
    u(t)=\sqrt{u_0^2+ \left(\frac{t-t_0}{t_E}\right)^2}
\end{equation}
where we have defined $u_0\equiv |\vec{w}_{L S, 0}|/\theta_E$
and $t_E\equiv \theta_E/|\boldsymbol{\mu}_{LS}|$.
The magnification is a function of three parameters $(t_0, u_0, t_E)$
and the resulting curve as a function of time is often called the Paczy\'nski curve
\citep{1986ApJ...304....1P,1986ApJ...301..503P}.
The magnification as a function of time for different impact parameters $u_0$
is shown in Figure~\ref{fig:paczynski_curve}. Notice the very steep fall off in
magnification as as we move away from the Einstein ring ($A(u)\propto
    1+2u^{-4}$).
In the limits of $u_0\rightarrow 0$ and $u_0\gg 1$  there is a continuous
mathematical degeneracy between the parameters $t_0$, $u_0$ and $t_E$
\citep{1997ApJ...487...55W}.

@@

\subsection{Observed flux}
The magnification $A(t)$ is not a direct observable in microlensing events. We
can only measure the flux of the source star which is usually contaminated with
the flux of other nearby stars in the dense stellar field and potentially also
with light from the lens and possible companions to the lens. The observed flux
is thus given by
\begin{equation}
    F(t)=F_S\,A(t)+F_B
    \label{eq:flux_observed_micro}
\end{equation}
where $F_s$ is the flux of the source star and $F_B$ is the contamination flux also
called the \emph{blending flux}. To quantify the strength of the blending e can
define the emph{source flux fraction} $b_S$ as the fraction of total (unmagnified)
flux that is coming from the source star:
\begin{equation}
    b_S\equiv F_S/(F_S + F_B)
\end{equation}
For highly blended events $b_S\approx 0$ and in absence of blending $b_S=1$.
The source flux $F_S$ and the blending flux $F_B$ are generally highly correlated
and it is often preferable to use a slightly different parametrization proposed by
\citet{2009MNRAS.393..816D}. In this parametrization we use the
\emph{baseline flux} $F_\mathrm{base}$ and the difference between
the baseline flux and the flux at peak magnification $\Delta F\equiv F_S[A(u_0)-1]$
as the new parameters.
Equation~\ref{eq:flux_observed_micro} then takes the form
\begin{equation}
    f=\Delta F\frac{A(t) - 1}{A(t_0)-1}+F_\mathrm{base}
    \label{eq:flux_dominik}
\end{equation}
It is generally straightforward to measure the flux difference $\Delta F$, the
baseline flux $F_\mathrm{base}$ and the center of the curve $t_0$ but getting a
good estimate of $t_E$ even outside of the regime $u_0\rightarrow 0$ and
$u_0\gg 1$ requires good photometric sampling along the ``wings'' of the light
curve \citep{2009MNRAS.393..816D}.

\subsection{Astrometric microlensing}
\label{ssec:astrometric_microlensing}
\begin{figure}[t]
    \begin{centering}
        \includegraphics[width=0.7\linewidth]{figures/single_lens_astrometric_shift.pdf}
        \caption{
            Astrometric shift in the centroid of light for a point source point lens
            model.
            The various curves correspond to different impact parameters $u_0$.}
        \label{fig:single_lens_astrometric_shift}
    \end{centering}
\end{figure}
